---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# RWisecondorX

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`RWisecondorX` exposes WisecondorX-style copy number analysis workflows from R on
top of `Rduckhts` and DuckDB.

The package is intentionally split into two layers:

- `bam_convert()` runs the convert step in R/SQL and does not require Python at runtime.
- `bam_convert_npz()` optionally uses `reticulate` plus `numpy` to write WisecondorX-compatible `.npz` files.
- `wisecondorx_newref()` and `wisecondorx_predict()` optionally call the official `wisecondorx` CLI through `condathis`.

The convert implementation is designed for exact conformance with upstream
WisecondorX duplicate handling, including the streaming `larp` / `larp2`
behaviour used during bin counting.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("sounkou-bioinfo/RWisecondorX")
```

`RWisecondorX` imports `Rduckhts`, `DBI`, and `duckdb`. Optional features use:

- `reticulate` for writing `.npz` files
- `condathis` for creating a reproducible conformance environment with the
  official bioconda `wisecondorx` package

## What The Package Covers

The current workflow surface is:

1. Convert BAM/CRAM alignments to WisecondorX-style per-bin counts with `bam_convert()`.
2. Optionally serialise those counts to `.npz` with `bam_convert_npz()`.
3. Optionally build a reference panel with `wisecondorx_newref()`.
4. Optionally run predictions with `wisecondorx_predict()`.

This keeps the core counting step in R while still allowing conformance checks
and downstream use of the official Python CLI when needed.

## Convert A BAM To Bin Counts

The package ships with a small chromosome 11 BAM fixture for examples and tests.

```{r convert-example}
library(RWisecondorX)

fixture_bam <- system.file(
  "extdata",
  "hg00106_chr11_fixture.bam",
  package = "RWisecondorX"
)

bins <- bam_convert(
  fixture_bam,
  binsize = 5000L,
  rmdup = "streaming"
)

names(Filter(Negate(is.null), bins))

length(bins[["11"]])

sum(bins[["11"]])
```

The `rmdup` argument supports the three intended duplicate-handling modes:

- `"streaming"`: exact replication of upstream WisecondorX `larp` / `larp2`
- `"none"`: no duplicate removal, corresponding to `wisecondorx convert --normdup`
- `"flag"`: use the SAM duplicate flag (`0x400`) for pre-marked BAMs

## Optional NPZ And CLI Workflow

When you want to continue into the original WisecondorX reference and prediction
steps, use the optional helpers below.

```r
# Requires reticulate + numpy
bam_convert_npz(
  bam = "sample.bam",
  npz = "sample.npz",
  binsize = 5000L,
  rmdup = "streaming"
)

# Requires condathis; installs official bioconda wisecondorx on first use
wisecondorx_newref(
  npz_files = c("control1.npz", "control2.npz"),
  output = "reference.npz",
  binsize = 5000L,
  ref_binsize = 50000L
)

wisecondorx_predict(
  npz = "sample.npz",
  ref = "reference.npz",
  output_prefix = "results/sample",
  ref_binsize = 50000L
)
```

## SRA Metadata Helpers

The package also includes small helpers for staging SRA run metadata used by
conformance fixtures.

```{r sra-example}
library(RWisecondorX)

sra_runinfo_url("PRJNA400134")
```

## Development

`README.Rmd` is the editable source for this document. Regenerate `README.md`
with the repository `Makefile`:

```sh
make readme
```
