---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# RWisecondorX

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`RWisecondorX` is an R port of [WisecondorX](https://github.com/CenterForMedicalGeneticsGhent/WisecondorX)
built on top of `Rduckhts` and DuckDB. It follows the
[rewrites.bio](https://rewrites.bio) principles: exact emulation of upstream
outputs, full credit to the original authors, and transparency about AI
assistance.

The core convert step runs entirely in R/SQL via `Rduckhts` with no Python
dependency. `bam_convert()` returns per-bin read counts in memory;
`bam_convert_bed()` writes them to a bgzipped, tabix-indexed BED file readable
by DuckDB or any tabix-aware tool; and `bam_convert_npz()` serialises them to a
WisecondorX-compatible `.npz` via `reticulate`. The convert implementation
exactly replicates the upstream `larp` / `larp2` streaming deduplication
behaviour, achieving bin-for-bin agreement with the Python implementation.

For pipelines that need the full upstream toolchain, thin `condathis` wrappers
cover every stage: `wisecondorx_convert()`, `wisecondorx_newref()`, and
`wisecondorx_predict()` each delegate to the official bioconda package without
requiring a manual conda setup.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("sounkou-bioinfo/RWisecondorX")
```

`RWisecondorX` imports `Rduckhts`, `DBI`, and `duckdb`. `reticulate` is needed
only to write `.npz` files; `condathis` is needed only for the upstream CLI
wrappers. Neither is required for the core convert step.

## Convert A BAM To Bin Counts

The package ships with a small chromosome 11 BAM fixture for examples and tests.

```{r convert-example}
library(RWisecondorX)

fixture_bam <- system.file(
  "extdata",
  "hg00106_chr11_fixture.bam",
  package = "RWisecondorX"
)

bins <- bam_convert(
  fixture_bam,
  binsize = 5000L,
  rmdup = "streaming"
)

names(Filter(Negate(is.null), bins))

length(bins[["11"]])

sum(bins[["11"]])
```

The `rmdup` argument controls duplicate handling. `"streaming"` exactly
replicates the upstream WisecondorX `larp` / `larp2` state machine and is the
default. `"none"` skips deduplication entirely, matching `wisecondorx convert
--normdup` for NIPT data where read depth is low. `"flag"` uses the SAM
duplicate flag (`0x400`) for BAMs already processed by Picard or sambamba.

## Optional BED.gz Output

`bam_convert_bed()` is the language-agnostic output path. It produces a
bgzipped BED file (`chrom`, `start`, `end`, `count`; 0-based coordinates) and a
tabix index alongside it, using `rduckhts_bgzip` and `rduckhts_tabix_index` from
`Rduckhts` — no external tools or Python required. The resulting file can be
queried by region directly from DuckDB.

```r
bam_convert_bed(
  bam     = "sample.bam",
  bed     = "sample.bed.gz",   # → sample.bed.gz + sample.bed.gz.tbi
  binsize = 5000L,
  rmdup   = "streaming"
)
```

## Optional NPZ And CLI Workflow

When the full upstream WisecondorX pipeline is needed, `wisecondorx_convert()`
wraps `wisecondorx convert` via `condathis` and produces an `.npz` through the
official Python implementation. `wisecondorx_newref()` and
`wisecondorx_predict()` cover the remaining stages. All three handle conda
environment creation automatically on first use.

```r
wisecondorx_convert(
  bam     = "sample.bam",
  npz     = "sample.npz",
  binsize = 5000L,
  normdup = FALSE          # TRUE → --normdup (NIPT)
)
```

```{r npz-example}
library(RWisecondorX)

fixture_npz <- tempfile(fileext = ".npz")
np <- reticulate::import("numpy", convert = FALSE)

bam_convert_npz(
  bam     = fixture_bam,
  npz     = fixture_npz,
  binsize = 5000L,
  rmdup   = "streaming",
  np      = np
)

c(
  npz_created = file.exists(fixture_npz),
  npz_size    = file.info(fixture_npz)$size
)
```

The CLI wrappers expose the upstream arguments documented in the WisecondorX
README. They delegate to the official `wisecondorx` bioconda package installed
automatically via `condathis`.

```{r cli-example, eval = FALSE}
library(RWisecondorX)

reference_npz <- tempfile(fileext = ".npz")
plotyfrac_png <- tempfile(fileext = ".png")
output_prefix <- tempfile(pattern = "wisecondorx-")

wisecondorx_newref(
  npz_files   = c(fixture_npz, fixture_npz),
  output      = reference_npz,
  binsize     = 5000L,
  ref_binsize = 50000L,
  nipt        = TRUE,
  refsize     = 300L,
  yfrac       = NULL,
  plotyfrac   = plotyfrac_png,
  cpus        = 1L
)

wisecondorx_predict(
  npz          = fixture_npz,
  ref          = reference_npz,
  output_prefix = output_prefix,
  ref_binsize  = 50000L,
  minrefbins   = 150L,
  maskrepeats  = 5L,
  zscore       = 5,
  alpha        = 1e-4,
  beta         = NULL,
  blacklist    = NULL,
  gender       = NULL,
  bed          = TRUE,
  plot         = FALSE,
  regions      = NULL,
  ylim         = NULL,
  cairo        = FALSE,
  seed         = 1L
)
```

## SRA Metadata Helpers

`sra_runinfo_url()`, `download_sra_runinfo()`, and `read_sra_runinfo()` are
small helpers for staging NCBI SRA run metadata used by conformance fixtures.

```{r sra-example}
library(RWisecondorX)

sra_runinfo_url("PRJNA400134")
```

## Development

`README.Rmd` is the editable source for this document. Run `make readme` to
rerender `README.md`, `make rd` to regenerate documentation from roxygen2
comments, and `make fixtures` to rebuild the bundled test data.
