---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(RWisecondorX)
```

# RWisecondorX

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/RWisecondorX/actions/workflows/R-CMD-check.yaml)
[![rewrites.bio](https://rewrites.bio/badges/rewrites-bio.svg)](https://rewrites.bio)
<!-- badges: end -->

`RWisecondorX` is an R toolkit for copy number analysis and trisomy prediction
in non-invasive prenatal testing (NIPT), built on top of `Rduckhts` and DuckDB.
It ports both [WisecondorX](https://github.com/CenterForMedicalGeneticsGhent/WisecondorX)
(CNV detection) and [NIPTeR](https://github.com/molgenis/NIPTeR) (trisomy
prediction via Z-scores, NCV, regression, and chi-squared correction), following
the [rewrites.bio](https://rewrites.bio) principles: exact emulation of upstream
outputs, full credit to the original authors, and transparency about AI assistance.

The core convert step now runs through the native `bam_bin_counts(...)` kernel
bundled in `Rduckhts`, with no Python dependency. `bam_convert()` returns
per-bin read counts in memory;
`bam_convert_bed()` writes them to a bgzipped, tabix-indexed BED file readable
by DuckDB or any tabix-aware tool; and `bam_convert_npz()` serialises them to a
WisecondorX-compatible `.npz` via `reticulate`. The convert implementation
exactly replicates the upstream `larp` / `larp2` streaming deduplication
behaviour, achieving bin-for-bin agreement with the Python implementation.

The NIPTeR statistical layer provides GC correction (via on-the-fly FASTA
computation, not bundled tables), chi-squared overdispersion correction,
chromosomal fraction Z-scores, normalised chromosome values (NCV), and forward
stepwise regression — all operating on `NIPTeRSample` objects produced by
`nipter_bin_bam()`.

For pipelines that need the full upstream toolchain, thin `condathis` wrappers
cover every stage: `wisecondorx_convert()`, `wisecondorx_newref()`, and
`wisecondorx_predict()` each delegate to the official bioconda package without
requiring a manual conda setup.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("sounkou-bioinfo/RWisecondorX")
```

`RWisecondorX` imports `Rduckhts`, `DBI`, and `duckdb`. `reticulate` is needed
only to write `.npz` files; `condathis` is needed only for the upstream CLI
wrappers. Neither is required for the core convert step.

## Convert A BAM To Bin Counts

The package ships with a small chromosome 11 BAM fixture for examples and tests.

```{r convert-example}
fixture_bam <- system.file(
  "extdata",
  "hg00106_chr11_fixture.bam",
  package = "RWisecondorX"
)

bins <- bam_convert(
  fixture_bam,
  binsize = 5000L,
  rmdup = "streaming"
)

names(Filter(Negate(is.null), bins))

length(bins[["11"]])

sum(bins[["11"]])
```

The `rmdup` argument controls duplicate handling. `"streaming"` exactly
replicates the upstream WisecondorX `larp` / `larp2` state machine and is the
default. `"none"` skips deduplication entirely, matching `wisecondorx convert
--normdup` for NIPT data where read depth is low. `"flag"` uses the SAM
duplicate flag (`0x400`) for BAMs already processed by Picard or sambamba.

## BED.gz Output And Round-Trip

`bam_convert_bed()` is the language-agnostic output path. It produces a
bgzipped BED file (`chrom`, `start`, `end`, `count`; 0-based coordinates) and a
tabix index, using `rduckhts_bgzip` and `rduckhts_tabix_index` from `Rduckhts`
— no external tools or Python required.

```{r bed-roundtrip}
bed_file <- tempfile(fileext = ".bed.gz")

bam_convert_bed(
  bam     = fixture_bam,
  bed     = bed_file,
  binsize = 5000L,
  rmdup   = "streaming"
)

c(
  bed_created   = file.exists(bed_file),
  index_created = file.exists(paste0(bed_file, ".tbi")),
  bed_size_kb   = round(file.info(bed_file)$size / 1024, 1)
)

# Round-trip: reload into the named-list format for rwisecondorx_newref()
sample_reloaded <- bed_to_sample(bed_file, binsize = 5000L)
length(sample_reloaded[["11"]])  # same as bins[["11"]] above
```

`bed_to_nipter_sample()` reads a 5-column NIPTeR BED.gz into a `NIPTeRSample`
object compatible with all NIPTeR statistical functions. Column count is
auto-detected (5 = CombinedStrands, 9 = SeparatedStrands).

## Native WisecondorX Pipeline (No Python)

`rwisecondorx_newref()` and `rwisecondorx_predict()` are pure R/Rcpp
implementations of the full WisecondorX algorithm — reference building, PCA
correction, within-sample normalisation, CBS segmentation, and aberration
calling — with no Python dependency. Performance-critical KNN reference-bin
finding is compiled via Rcpp with OpenMP parallelisation. CBS segmentation uses
DNAcopy (or ParDNAcopy for parallel operation) directly.

The example below generates a 50-sample synthetic cohort (35 euploid female,
12 euploid male, 1×T21, 1×T18, 1×T13), builds a reference, and verifies that
a trisomy sample is correctly detected.

```{r cohort-example}
library(DNAcopy)

# generate_cohort() uses compressed chromosome lengths (100bp = 1 genomic bin)
# so each BAM is ~435KB. Requires samtools on PATH.
has_samtools <- nchar(Sys.which("samtools")) > 0L
if (!has_samtools) {
  message("samtools not found — skipping cohort pipeline example")
  knitr::opts_chunk$set(eval = FALSE)
}
```

```{r cohort-run, eval = has_samtools}
cohort_dir <- file.path(tempdir(), "rwx_readme_cohort")
manifest   <- suppressMessages(generate_cohort(cohort_dir, verbose = FALSE))

# Cohort composition
table(manifest$sex, manifest$trisomy)
```

```{r cohort-bin, eval = has_samtools}
# Bin all 50 BAMs
all_samples <- lapply(
  file.path(cohort_dir, manifest$bam_file),
  function(bam) suppressMessages(
    bam_convert(bam, binsize = COMPRESSED_BINSIZE, mapq = 0L, rmdup = "none")
  )
)
names(all_samples) <- manifest$sample_id
```

```{r cohort-newref, eval = has_samtools}
# Build reference (refsize=10 for speed; production uses refsize=300)
ref <- suppressMessages(
  rwisecondorx_newref(
    samples = all_samples,
    binsize = COMPRESSED_BINSIZE,
    nipt    = TRUE,
    refsize = 10L,
    cpus    = 1L
  )
)
c(binsize = ref$binsize, masked_bins = sum(!ref$mask), is_nipt = ref$is_nipt)
```

```{r cohort-predict, eval = has_samtools}
# Predict on the T21 sample
t21_id   <- manifest$sample_id[manifest$trisomy == "T21"]
pred_t21 <- suppressMessages(
  rwisecondorx_predict(
    sample    = all_samples[[t21_id]],
    reference = ref,
    zscore    = 3,
    alpha     = 1e-4,
    seed      = 42L
  )
)

# Chr21 should show gain
pred_t21$aberrations[pred_t21$aberrations$chr == "21", ]
```

```{r cohort-cleanup, eval = has_samtools, include = FALSE}
unlink(cohort_dir, recursive = TRUE)
```

### Multi-File BED Input For Reference Building

`rwisecondorx_newref()` also accepts a directory of 4-column BED.gz files
produced by `bam_convert_bed()` via the `bed_dir` parameter. All matching files
are loaded in a single DuckDB pass — no in-memory sample list needed:

```r
ref <- rwisecondorx_newref(
  bed_dir = "bed_counts/",     # directory of *.bed.gz files
  binsize = 100000L,
  nipt    = TRUE,
  cpus    = 4L
)
```

## Optional NPZ And CLI Workflow

When the full upstream WisecondorX pipeline is needed, `wisecondorx_convert()`
wraps `wisecondorx convert` via `condathis` and produces an `.npz` through the
official Python implementation. All three wrappers handle conda environment
creation automatically on first use.

```r
wisecondorx_convert(
  bam     = "sample.bam",
  npz     = "sample.npz",
  binsize = 5000L,
  normdup = FALSE          # TRUE → --normdup (NIPT)
)
```

```{r npz-example, eval = requireNamespace("reticulate", quietly = TRUE)}
fixture_npz <- tempfile(fileext = ".npz")
np <- reticulate::import("numpy", convert = FALSE)

bam_convert_npz(
  bam     = fixture_bam,
  npz     = fixture_npz,
  binsize = 5000L,
  rmdup   = "streaming",
  np      = np
)

c(
  npz_created = file.exists(fixture_npz),
  npz_size    = file.info(fixture_npz)$size
)
```

The CLI wrappers expose the upstream arguments documented in the WisecondorX
README and delegate to the official `wisecondorx` bioconda package.

```{r cli-example, eval = FALSE}
# Requires: condathis installed; wisecondorx bioconda package auto-installed

npz_files <- list.files("controls/", "\\.npz$", full.names = TRUE)

wisecondorx_newref(
  npz_files   = npz_files,
  output      = "reference.npz",
  ref_binsize = 100000L,
  nipt        = TRUE,
  refsize     = 300L,
  cpus        = 4L
)

wisecondorx_predict(
  npz           = "sample.npz",
  ref           = "reference.npz",
  output_prefix = "results/sample",
  bed           = TRUE,
  seed          = 1L
)
```

## Beta Mode (Somatic CNV)

`rwisecondorx_predict()` supports a `beta` parameter for purity-based
aberration calling, appropriate for somatic CNV analysis where tumour purity is
known. When `beta` is supplied, ratio cutoffs replace Z-score thresholds:

```
gain cutoff: log2((ploidy + beta/2) / ploidy)
loss cutoff: log2((ploidy - beta/2) / ploidy)
```

`beta` is the tumour purity estimate (0–1; 0 = most liberal calling,
1 = most conservative; optimally close to true purity). This mode is **not
appropriate for NIPT**, which uses the default Z-score calling.

```r
# Somatic CNV: 40% purity tumour sample
pred_somatic <- rwisecondorx_predict(
  sample    = tumour_sample,
  reference = ref,
  beta      = 0.4,   # ~tumour purity
  seed      = 1L
)
```

## Synthetic Cohort For Testing

`generate_cohort()` creates a 50-sample synthetic BAM cohort using compressed
chromosome lengths (100bp = 1 genomic 100kb bin), so each BAM is ~435KB. The
cohort contains euploid females and males plus three trisomy samples (T21, T18,
T13). See the native pipeline section above for end-to-end usage.

```{r cohort-example-minimal, eval = FALSE}
out_dir <- tempfile("cohort_")
manifest <- generate_cohort(out_dir)

# Bin with COMPRESSED_BINSIZE (100L) to match synthetic coordinates
bams    <- file.path(out_dir, manifest$bam_file)
samples <- lapply(bams, bam_convert, binsize = COMPRESSED_BINSIZE, rmdup = "none")
```

## NIPTeR-Style Trisomy Prediction

`RWisecondorX` ports the statistical analysis layer from
[NIPTeR](https://github.com/molgenis/NIPTeR) (de Weerd et al.), replacing
`Rsamtools` with `Rduckhts` and bundled `sysdata.rda` tables with on-the-fly
computation via `rduckhts_fasta_nuc()`. The analysis pipeline follows NIPTeR's
workflow: bin samples, build a control group, correct for GC bias and
overdispersion, then compute Z-scores or NCV values for trisomy prediction.

### Binning

`nipter_bin_bam()` produces `NIPTeRSample` objects compatible with all
downstream functions. Pre-filtering flags mirror `samtools view -f`/`-F`
conventions, so real-world NIPT pipelines that mark duplicates with Picard can
pass `exclude_flags = 1024L` directly.

```r
samples <- lapply(bam_files, nipter_bin_bam, binsize = 50000L)
```

### Control Group Construction

A control group is the reference cohort against which test samples are scored.
`nipter_as_control_group()` validates and deduplicates the input samples.
`nipter_diagnose_control_group()` reports per-chromosome Z-scores and
Shapiro-Wilk normality p-values across the group, useful for identifying
outlier samples before scoring. When working with a large cohort,
`nipter_match_control_group()` selects the closest-matching subset for a given
test sample based on chromosomal fraction distance (Rcpp+OpenMP accelerated).
`nipter_match_matrix()` computes the full N×N pairwise SSD matrix — useful for
QC outlier detection before building the control group.

```r
cg <- nipter_as_control_group(samples)
diag <- nipter_diagnose_control_group(cg)
matched_cg <- nipter_match_control_group(test_sample, cg, n = 50L)
mm <- nipter_match_matrix(cg, cpus = 4L)
```

### GC Correction

`nipter_gc_correct()` adjusts bin counts for GC-content bias. GC percentages
are computed per bin from the reference FASTA via `rduckhts_fasta_nuc()` rather
than from static bundled tables. Two methods are available: LOESS regression
(default, matching NIPTeR) and bin-weight normalisation. For cohort workflows,
`nipter_gc_precompute()` pre-computes and saves GC fractions to a TSV.bgz file
so the FASTA is only scanned once.

```r
# Pre-compute GC table (one-time per genome / binsize)
nipter_gc_precompute(fasta = "hg38.fa", binsize = 50000L,
                     out = "gc_table.tsv.bgz")

# Apply GC correction using pre-computed table (fast)
cg          <- nipter_gc_correct(cg, gc_table = "gc_table.tsv.bgz")
test_sample <- nipter_gc_correct(test_sample, gc_table = "gc_table.tsv.bgz")
```

### Chi-Squared Correction

`nipter_chi_correct()` identifies overdispersed bins across the control group
using a normalised chi-squared test and downweights them. The correction is
applied simultaneously to both the test sample and control group, maintaining
consistency for downstream scoring.

```r
corrected   <- nipter_chi_correct(test_sample, cg, chi_cutoff = 3.5)
test_sample <- corrected$sample
cg          <- corrected$control_group
```

### Scoring

`nipter_z_score()` computes the chromosomal fraction Z-score for a focus
chromosome. A Z-score above 3 is the conventional threshold for trisomy.

```r
z21 <- nipter_z_score(test_sample, cg, chromo_focus = 21)
z21$sample_z_score
z21$control_statistics
```

`nipter_ncv_score()` computes the normalised chromosome value, searching for
the denominator chromosome set that minimises control-group variance before
computing the Z-score. More robust than the standard Z-score when the control
group is small.

```r
ncv21 <- nipter_ncv_score(test_sample, cg, chromo_focus = 21)
ncv21$sample_ncv_score
ncv21$denominators
```

`nipter_regression()` uses forward stepwise linear regression to predict the
focus chromosome's fraction from other chromosomes, then scores the residual.

```r
reg21 <- nipter_regression(test_sample, cg, chromo_focus = 21)
reg21$models[[1]]$z_score
reg21$models[[1]]$predictors
```

## Sex Prediction

`nipter_sex_model()` trains a 2-component Gaussian mixture model on chromosome
fraction data from a control group for sex prediction. Three model types are
available: Y-fraction alone, bivariate X+Y fractions, and Y-unique region
read ratios from specific Y-chromosome genes. `nipter_predict_sex()` classifies
a test sample using one or more models with majority-vote consensus
(ties → female, the conservative choice for NIPT).

```{r sex-prediction, eval = FALSE}
# Requires: mclust package; whole-genome BAMs; FASTA for GC correction
cg <- nipter_as_control_group(samples)
cg <- nipter_gc_correct(cg, fasta = "hg38.fa")
corrected <- nipter_chi_correct(samples[[1]], cg)
cg <- corrected$control_group

model_y  <- nipter_sex_model(cg, method = "y_fraction")
model_xy <- nipter_sex_model(cg, method = "xy_fraction")

nipter_predict_sex(test_sample, models = list(model_y, model_xy))
```

## SRA Metadata Helpers

`sra_runinfo_url()`, `download_sra_runinfo()`, and `read_sra_runinfo()` are
small helpers for staging NCBI SRA run metadata used by conformance fixtures.

```{r sra-example}
sra_runinfo_url("PRJNA400134")
```

## Validation And Conformance

Per [rewrites.bio](https://rewrites.bio) principle 4.3, see
[`VALIDATION.md`](VALIDATION.md) for a full account of what has been validated,
on what data, against which upstream versions, and what remains unvalidated.

**Summary of current status:**

- `bam_convert()`: bin-for-bin identical to Python WisecondorX on
  `HG00106.chrom11` (0 mismatches, 25,115 bins).
- NIPTeR statistical layer (Z-score, NCV, chi, regression): formulas verified
  against inline reference implementations in `test_nipter_conformance.R`.
- Trisomy detection: T21/T18/T13 all correctly called in
  `test_cohort_pipeline.R` and `test_wisecondorx_e2e.R` using synthetic cohort.
- `rwisecondorx_newref/predict` segment and aberration comparison against
  upstream Python: **pending real-data run** (run
  `inst/scripts/real_data_conformance.R` with your server BAMs).
- Beta mode (purity-based calling): implemented, not yet tested with clinical
  samples.

## Development

`README.Rmd` is the editable source for this document. Run `make readme` to
rerender `README.md`, `make rd` to regenerate documentation from roxygen2
comments, `make fixtures` to rebuild the bundled test data, and
`make conformance` to run the end-to-end WisecondorX conformance test.
